---
title: "Final Project"
author: "Samit Kafle"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
---

**Turn off the warning in markdown**
```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


**Automatically install and load required libraries**
```{r package-setup, include=FALSE}
required_packages <- c(
  "tidyverse", "ggplot2", "lme4", "car", "gridExtra", "corrplot",
  "BlandAltmanLeh", "DescTools", "knitr", "rmarkdown", "ggpmisc", "ggpubr"
)

# Install any missing packages
installed_packages <- rownames(installed.packages())
for (pkg in required_packages) {
  if (!(pkg %in% installed_packages)) {
    install.packages(pkg, dependencies = TRUE)
  }
}

# Load packages
lapply(required_packages, library, character.only = TRUE)
```




**Read the data**
```{r}
data <- read.csv("TLSvsiTree/tlsitree.csv")
```

**Ensuring proper data structure and calculating the difference in estimates between two methods**
```{r}
clean_data <- data %>%
  drop_na() %>%
  mutate(
    difference = carbon_tls - carbon_itree,
    percent_diff = (difference/carbon_itree) * 100
  )

```

**Descriptive Statistics**
```{r}
##Summary statistics of individual species
summary_stats <- clean_data %>%
  group_by(species) %>%
  summarise(
    mean_tls = mean(carbon_tls),
    sd_tls = sd(carbon_tls),
    mean_itree = mean(carbon_itree),
    sd_itree = sd(carbon_itree),
    mean_diff = mean(difference),
    rmse = sqrt(mean(difference^2)),
    ccc = CCC(carbon_tls, carbon_itree)$rho.c,
    pearson_r = cor(carbon_tls, carbon_itree, method = "pearson"),
    pearson_p_value = cor.test(carbon_tls, carbon_itree, method = "pearson")$p.value
  )
##Summary statistics of all species combined
summary_stats_combined <- clean_data %>%
  summarise(
    mean_tls = mean(carbon_tls),
    sd_tls = sd(carbon_tls),
    mean_itree = mean(carbon_itree),
    sd_itree = sd(carbon_itree),
    mean_diff = mean(difference),
    rmse = sqrt(mean(difference^2)),
    ccc = CCC(carbon_tls, carbon_itree)$rho.c,
    pearson_r = cor(carbon_tls, carbon_itree, method = "pearson"),
    pearson_p_value = cor.test(carbon_tls, carbon_itree, method = "pearson")$p.value
  )
  
```

### Summary Statistics of carbon estimates
```{r}
summary_stats_combined$species="combined"
descriptive_stats=rbind(summary_stats,summary_stats_combined)
kable(descriptive_stats,digits = 2)
```


**Test for Normality using shapiro wilk normality test**
```{r}
model=lm(carbon_tls~carbon_itree,data=clean_data)
shapiro.test(resid(model))
```

we got p<0.05. Hence the data is not normally distributed. So we used Wilcoxon paired t test to compare the difference in mean between itree and tls based carbon 

```{r}
wilcox_test=wilcox.test(clean_data$carbon_itree,clean_data$carbon_tls, paired = TRUE)
wilcox_test
```

**Subsetting only values to compare two methods for all species combined**

```{r}
combined_data=clean_data[,2:3]
colnames(combined_data)=c("tls","itree")
```

### Box plot for all species combined using wilcox paired t test
```{r}

ggplot(pivot_longer(combined_data, cols = everything(), 
                    names_to = "method", values_to = "carbon_kg"), ##converting into long format
       aes(x = method, y = carbon_kg, fill = method)) +
  geom_boxplot(fill=c("pink","skyblue")) +
  stat_compare_means(label = "p.format",col="darkgreen",paired=T)+
  theme_classic() +
  labs(title = "Comparision between two methods of carbon estimates",
       x = "Method", y = "Carbon Estimate (kg)") +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```

### Box plots by species (Individual plots per species) using wilcox t test

```{r}
ggplot(clean_data %>%
  group_by(TreeID, species) %>%
  pivot_longer(cols = c(carbon_tls, carbon_itree), 
               names_to = "type", values_to = "carbon"), aes(x = type, y = carbon, fill = type)) +
  geom_boxplot(fill=c("pink","skyblue","pink","skyblue","pink","skyblue")) +
  stat_compare_means(paired=T, label = "p.format",col="darkgreen")+
  facet_wrap(~ species, scales = "free_y") +
  scale_x_discrete(labels = c("carbon_itree" = "itree", "carbon_tls" = "tls"))+
  labs(title = "Species wise comparision",
       x = "Method",
       y = "Carbon (kg)") +
  theme_classic()+theme(axis.text.x = element_text(angle = 0, hjust = 0.5),strip.background=element_blank())
```

### Combined species Scatter plot with regression line
```{r}
scatter_plot_combined <- ggplot(clean_data, aes(x = carbon_itree, y = carbon_tls)) +
  geom_point(aes(color = species), alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_smooth(method = "lm", se = TRUE) +
  stat_poly_eq(aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")), 
                                                      formula = y ~ x, 
                                                      parse = TRUE, 
                                                      size = 4)+
  theme_classic() +
  labs(x = "i-Tree Carbon Estimate (kg)",
       y = "TLS Carbon Estimate (kg)",
       title = "Comparison of Carbon Estimates between Methods")
scatter_plot_combined
```

### Scatter plot with regression lines for each species
```{r}
scatter_plot_combined <- ggplot(clean_data, aes(x = carbon_itree, y = carbon_tls)) +
  geom_point( alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_smooth(method = "lm", se = TRUE) +
  stat_poly_eq(aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")), 
                                                      formula = y ~ x, 
                                                      parse = TRUE, 
                                                      size = 3)+
  facet_wrap(~species)+
  theme(legend.position = "none") +
  theme_classic() +
  labs(x = "i-Tree Carbon Estimate (kg)",
       y = "TLS Carbon Estimate (kg)",
       title = "Comparison of Carbon Estimates between Methods"
       )
scatter_plot_combined
```


### Create scatter plot with regression lines by species and stacked R² values
```{r}

# Calculate regression statistics for each species
species_stats <- clean_data %>%
  group_by(species) %>%
  summarise(
    r2 = summary(lm(carbon_tls ~ carbon_itree))$r.squared,
    intercept = coef(lm(carbon_tls ~ carbon_itree))[1],
    slope = coef(lm(carbon_tls ~ carbon_itree))[2]
  )

# Create positions for stacked R² values
species_stats <- species_stats %>%
  mutate(
    y_position = seq(
      from = max(clean_data$carbon_tls),
      by = -max(clean_data$carbon_tls)/20,  # Adjust this divisor to change spacing
      length.out = n()
    )
  )

# Create the plot
scatter_plot <- ggplot(clean_data, aes(x = carbon_itree, y = carbon_tls, color = species)) +
  # Add points
  geom_point(alpha = 0.6) +
  # Add 1:1 reference line
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  # Add regression lines by species
  geom_smooth(method = "lm", se = TRUE, aes(fill = species)) +
  # Add R² values in stacked format
  geom_text(
    data = species_stats,
    aes(
      x = min(clean_data$carbon_itree),
      y = y_position,
      label = sprintf("%s: R² = %.3f", species, r2),
      color = species
    ),
    hjust = 0,
    vjust = 0,
    size = 3,
    fontface = "bold"
  ) +
  # Customize theme and labels
  theme_classic() +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  ) +
  labs(
    x = "i-Tree Carbon Estimate (kg)",
    y = "TLS Carbon Estimate (kg)",
    title = "Comparison of Carbon Estimates between Methods",
    subtitle = "Regression lines shown by species",
    color = "Tree Species",
    fill = "Tree Species"
  )
```



### Regression statistics
```{r}
kable(data.frame(species_stats))
scatter_plot
```


**Mainstem vs Branches and branchorder**
**read the data**

```{r}
data2=read.csv("branchorder/combined_tree_data.csv")

```

**Remove branch_order = 0 (main stem)**
```{r}
branches_only <- data2 %>% filter(branch_order != 0)
```

**Create a summary for each tree with main stem and branch carbon**
```{r}
carbon_summary <- data2 %>%
  group_by(TreeID) %>%
  summarize(
    main_stem_carbon = sum(carbon[branch_order == 0], na.rm = TRUE),
    branch_carbon = sum(carbon[branch_order != 0], na.rm = TRUE)
  )
```



### Combined Carbon by Branch Order (All species combined)
```{r}
ggplot(branches_only, aes(x = factor(branch_order), y = carbon, fill = factor(branch_order))) +
  geom_boxplot(fill="skyblue") +
  theme_classic() +
  stat_compare_means(label = "p.format",col="darkgreen")+
  labs(title = "Combined Plot: Carbon by Branch Order (All Species)",
       x = "Branch Order",
       y = "Carbon (kg)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

###  Box plot of carbon by branch order for each species
```{r}
ggplot(branches_only, aes(x = factor(branch_order), y = carbon, fill = factor(branch_order))) +
  geom_boxplot(fill="skyblue") +
  stat_compare_means(label = "p.format",col="darkgreen")+
  facet_wrap(~ Species, scales = "free_y") +
  labs(title = "Box Plot of Carbon by Branch Order per Species",
       x = "Branch Order",
       y = "Carbon") +
  theme_classic()+theme(strip.background = element_blank())
```

### Summary for main stem and branch carbon
```{r}
carbon_summary <- data2 %>%
  group_by(TreeID, Species) %>%
  summarize(
    main_stem_carbon = sum(carbon[branch_order == 0], na.rm = TRUE),# 0 order branch is mainstem
    branch_carbon = sum(carbon[branch_order != 0], na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(main_stem_carbon, branch_carbon), names_to = "type", values_to = "carbon")
```



### Main Stem vs Branch Carbon (Individual plots per species) 
```{r}
ggplot(carbon_summary, aes(x = type, y = carbon, fill = type)) +
  geom_boxplot(fill=c("pink","skyblue","pink","skyblue","pink","skyblue")) +
  stat_compare_means( label = "p.format",col="darkgreen")+
  facet_wrap(~ Species, scales = "free_y") +
  scale_x_discrete(labels = c("main_stem_carbon" = "Main Stem", "branch_carbon" = "Branch"))+
  labs(title = "Main Stem vs Branch Carbon (Individual Plots per Species)",
       x = "Type (Main Stem vs Branch)",
       y = "Carbon (kg)") +
  theme_classic()+theme(axis.text.x = element_text(angle = 0, hjust = 0.5),strip.background = element_blank())
```


### Combined Main Stem vs Branch Carbon (All species combined)
```{r}
ggplot(carbon_summary, aes(x = type, y = carbon, fill = type)) +
  geom_boxplot(fill=c("pink","skyblue")) +
  stat_compare_means( label = "p.format",col="darkgreen")+
  theme_classic() +
  scale_x_discrete(labels = c("main_stem_carbon" = "Main Stem", "branch_carbon" = "Branch"))+
  labs(title = "Combined Plot: Main Stem vs Branch Carbon (All Species)",
       x = "Type",
       y = "Carbon (kg)") +
  theme_classic()
```

### Main Stem to Branch Carbon Ratio group by species

```{r}
carbon_ratio <- carbon_summary %>%
  group_by(Species) %>%
  summarise(
    Main_Stem_Carbon = mean(carbon[type == "main_stem_carbon"], na.rm = TRUE),
    Branch_Carbon = mean(carbon[type == "branch_carbon"], na.rm = TRUE),
    Ratio = Main_Stem_Carbon / Branch_Carbon  # Compute the ratio
    )
```


```{r}
kable(carbon_ratio,digits = 3)
```

